# Human Activity Prediction
========================================================

## Introduction

Use practical machine learning to predict human activity on HAR dataset. The dataset has been obtained from http://groupware.les.inf.puc-rio.br/har. 


## Pre-Processing data
The labelled data consists of a CSV file which can be loaded in R. Pre-processing includes all the steps to be performed before we actually start applying algorithms to learn from data. It includes generating train/test sets, cleanup, and normalization.

### Generate train / test sets
We generate train and test sets out of the original labelled data for evaluating our model.

```{r cache=TRUE, results='hide'}
library(caret);

#Load data, treat empty and #DIV/0! strings as NA -
dataCSV <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"));

# Set seed for reproducability
set.seed(13309);

# Partition to keep 80% data for training, and 20% for testing/cross-validation
inTrain <- createDataPartition(y=dataCSV$classe, p=0.8, list=FALSE)
trainCSV <- dataCSV[inTrain,]
testCSV <- dataCSV[-inTrain,]
```

So data is divided from total `r nrow(dataCSV)` to training set with `r nrow(trainCSV)` rows and test set with `r nrow(testCSV)` rows. We will not touch the test set, treat it as a blind data, and perform the rest of pre-processing by exploring training data. 

### Cleaning data
Training set has `r ncol(trainCSV)` columns. We reduce them as follows -

*Remove un-predictive columns*
The first column 'X' is just an index column, so we remove it.

```{r cache=TRUE, results='hide'}
trainClean1 <- trainCSV[,-1]
```

*Remove empty columns*
We observe that a lot of columns consist only of NA in data -

```{r fig.width=7, fig.height=6}
plot(colSums(is.na(trainClean1)),ylab="Count of NAs",main="NA frequency per column in training data")
```

```{r cache=TRUE, results='hide'}
# 95% of row count -
nrow_95_pc <- (9.5*nrow(trainClean1))/10

# Columns with more than 95% rows as NA
columns_with_95_pc_nas <- which(colSums(is.na(trainClean1))>nrow_95_pc)
```

There are `r length(columns_with_95_pc_nas)` columns with more than 95% rows as NA. So we remove those columns as part of cleanup -

```{r cache=TRUE, results='hide'}
trainClean2 <- trainClean1[,-columns_with_95_pc_nas]
```

*Reduce correlated numeric variables*
We identify the pair of numeric variables with high correlation

```{r cache=TRUE, results='hide'}
# Columns of type numeric -
numeric_columns <- which(sapply(trainClean2,class)!="factor")

count_numeric <- length(numeric_columns)

# Find index pairs with absolute correlation > 0.8
all_correlated_flat_pairs <- 
  which(abs(cor(trainClean2[numeric_columns],trainClean2[numeric_columns]))>0.8)

# Remove the pairs like (1,1) (2,2) etc.
same_flat_pairs <- sapply(c(0,seq_along(numeric_columns)),function(i){count_numeric*i + i + 1})
correlated_flat_pairs <- setdiff(all_correlated_flat_pairs, same_flat_pairs)

# This is a list of unique correlated index pairs,
correlated_index_pairs <- unique(lapply(correlated_flat_pairs,function(i){
  index1 <- floor(i/count_numeric)+1;
  index2 <- i%%count_numeric;
  if(index2 < index1) {
    temp <- index2;
    index2 <- index1;
    index1 <- temp;
  }
  c(numeric_columns[index1],numeric_columns[index2])
  }))
```

Let's take the first 2 pairs with high correlation -
```{r cache=TRUE}
correlated_index_pairs[1:2]
```

*Plotting corrrelated columns*
As we see, roll_belt (index 7), yaw_belt(index 9) and total_accel_belt(index 10) have high mutual corrrelation. Here is a plot depicting the same, darker shade implies higher correlation.

```{r fig.width=7, fig.height=6}
library(corrgram)

#Plot correlation for columns : 7,8,9,10
corrgram(trainClean2[,7:10],panel=panel.shade)
```

So we remove the first index of each such pairs.
```{r cache=TRUE, results='hide'}
remove_numeric_columns <- unique(sapply(correlated_index_pairs,function(elem){elem[[1]]}))

trainCleanFinal <- trainClean2[,-remove_numeric_columns]
```

To put things in perspective, we notice that shades are not that dark now -

```{r fig.width=7, fig.height=6}
#Plot correlation for columns : 7,8,9,10
corrgram(trainCleanFinal[,7:10],panel=panel.shade)
```


## Modelling
We train a random forest algorithm over the cleaned training set.

```{r cache=TRUE, results='hide'}
# Use parallelization across cores for quicker computation
library(doParallel)
cl <- makeCluster(detectCores())
registerDoParallel(cl)

# Train a random forest (rf) model with standardized training set and using cross validation
trainCleanFinal$classe <- factor(trainCleanFinal$classe)
rf_model <- train(classe ~ ., 
                  data=trainCleanFinal, 
                  method="rf", 
                  preProcess = c("center", "scale"), 
                  trControl=trainControl(method="cv"))
print(rf_model)
```

```{r cache=TRUE}
print(rf_model)
```

## Cross validation
We now analyze our model by computing in-sample error (i.e. error rate on training set) and out of sample error (i.e. error rate on testing set).

### In sample error
```{r cache=TRUE, results='hide'}
train_total_rows <- nrow(trainCleanFinal)

train_prediction <- predict(rf_model, trainCleanFinal)

train_correctly_predicted_rows <- length(which(train_prediction==trainCleanFinal$classe))
in_sample_error <- 100 * (1.0-(1.0*train_correctly_predicted_rows/train_total_rows))
```

```{r cache=TRUE}
table(train_prediction, trainCleanFinal$classe)
```
The in-sample error rate for the learnt model is `r in_sample_error`%.

### Out of sample error
```{r cache=TRUE, results='hide'}
testClean1 <- testCSV[,-1]
testClean2 <- testClean1[,-columns_with_95_pc_nas]
testCleanFinal <- testClean2[,-remove_numeric_columns]
test_total_rows <- nrow(testCleanFinal)

test_prediction <- predict(rf_model, testCleanFinal)

test_correctly_predicted_rows <- length(which(test_prediction==testCleanFinal$classe))
out_of_sample_error <- 100 * (1.0-(1.0*test_correctly_predicted_rows/test_total_rows))
```

```{r cache=TRUE}
table(test_prediction, testCleanFinal$classe)
```
The out-of-sample error rate for the learnt model is `r out_of_sample_error`%.
